---
title: "PatientLevelPrediction with Strategus"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
  error: true
---

# Overview

::: {.panel-tabset}

### Goals
This document shows how to:

* Load cohort JSONs and generate cohorts using `CohortGenerator`.
* Configure a `PatientLevelPrediction` (PLP) design.
* Assemble `Strategus` analysis specifications
* Execute the analysis specifications against a DuckDB Synthea example database.

### Workflow
```{mermaid}
flowchart LR
    A[Cohort JSONs] --> B(Cohort Generator)
    B --> C(Shared Resources)
    C --> D(Strategus Specs)
    D --> E(Execution Settings)
    E --> F[Results Review]
```

:::

This is the tutorial reference. To run the tutorial we recommend using the R file [here](https://github.com/OHDSI/Tutorial-PLP/blob/tutorial-code/exampleCode/networkStudy.R) which contains the same code but is easier to run and modify step-by-step.


# Setup

::: {.panel-tabset}

### Steps
Here we need to set up our environment.

::: {.callout-tip}
## Prerequisites
- [ ] If you don't have R installed please follow the HADES installation instructions for your operating system from [here](https://ohdsi.github.io/Hades/articles/rSetup.html).
- [ ] Usually to run OHDSI packages you need to install at least R, java and a github account (to install packages from github without rate-limiting).
- [ ] Once you have that set up you can install the required packages for this tutorial. They are `PatientLevelPrediction`, `CohortGenerator`, `CirceR`, and `Strategus`.
- [ ] Also to use the example database you need to install `duckdb`. Since `Strategus` is still under development you need to install it from github which requires the `remotes` package.
- [ ] Further for viewing the results you will need `OHDSIShinyAppBuilder` and `OhdsiShinyModules` the latter on github.
:::

::: {.callout-note}
## Example Data
Further please download the example database from [here](https://huggingface.co/datasets/egillax/OHDSI-ExampleData/resolve/main/database-1M_filtered.duckdb) and place in your location of choice. Then update the `dbPath` variable to point to that location.
:::

### R Code
```{r}
#| label: setup
# Install required packages (uncomment if needed):
# install.packages(c("PatientLevelPrediction", "CohortGenerator", "CirceR", "remotes", "duckdb", "OHDSIShinyAppBuilder"))
# remotes::install_github("OHDSI/Strategus")
# remotes::install_github("OHDSI/OhdsiShinyModules"))

library(PatientLevelPrediction)
library(Strategus)

seed <- 42
```

```{r}
dbPath <- "~/database/database-1M_filtered.duckdb"
```

```{r}
cdmSchema <- "main"
workSchema <- "cohorts"
cohortTable <- "cohort_table"
```

:::

# Cohort Generation

::: {.panel-tabset}

### Steps
To generate the cohorts we need to load the JSON files and create a cohort definition set. Then we can create the shared resource specifications and module specifications for cohort generation.

::: {.callout-warning}
## Cohort JSONs
- [ ] Download the JSON files from [here](https://github.com/OHDSI/Tutorial-PLP/tree/tutorial-code/exampleCode/cohorts).
- [ ] Open each file and press download.
- [ ] Save them somewhere locally and update the paths below to point to the location where you saved them.
:::

### R Code
```{r} 
#| label: cohort-generation
targetFile <- "./cohorts/31_[T] Pandemic Prediction new target.json"
outcomeFile <- "./cohorts/14_[O] Pandemic Prediction outcome pneumonia in hospital.json"

cohortDefinitions <- CohortGenerator::createEmptyCohortDefinitionSet()
cohortDefinitions[1, ] <- list(
  cohortId = 1,
  cohortName = "Target",
  sql = "",
  json = readChar(targetFile, file.info(targetFile)$size)
)
cohortDefinitions[2, ] <- list(
  cohortId = 2,
  cohortName = "Outcome",
  sql = "empty",
  json = readChar(outcomeFile, file.info(outcomeFile)$size)
)

cohortGeneratorModule <- CohortGeneratorModule$new()
cohortDefShared <- cohortGeneratorModule$createCohortSharedResourceSpecifications(cohortDefinitions)

cohortGeneratorModuleSpecifications <- cohortGeneratorModule$createModuleSpecifications(
  generateStats = TRUE
)
```

:::

# PLP Design

::: {.panel-tabset}

### Steps
This section configures the PLP design by defining covariates, population settings, preprocessing, data splits, and the baseline model choice.

::: {.callout-tip}
## Model Variations
- [ ] Use the baseline `setLassoLogisticRegression` call to reproduce the tutorial results.
- [ ] Swap to the alternative model options listed in the code comments when you are ready to experiment.
:::

### R Code
```{r}
#| label: plp-design

covariateSettings <- FeatureExtraction::createDefaultCovariateSettings()

populationSettings <- createStudyPopulationSettings(
  binary = TRUE,
  removeSubjectsWithPriorOutcome = TRUE,
  requireTimeAtRisk = FALSE,
  riskWindowStart = 1,
  riskWindowEnd = 30,
)

preprocessSettings <- createPreprocessSettings(
  minFraction = 0.001,
  normalize = TRUE,
  removeRedundancy = TRUE
)

splitSettings <- createDefaultSplitSetting(splitSeed = seed)

# Choose model based on parameter (baseline: lasso logistic)
modelSettings <- PatientLevelPrediction::setLassoLogisticRegression(seed = seed)
  # Other options (require additional packages; see Exercises below):
  # PatientLevelPrediction::setGradientBoostingMachine(ntrees = 100, seed = seed),
  # PatientLevelPrediction::setRandomForest(mtries = 10, seed = seed),
  # PatientLevelPrediction::setXgBoost(nrounds = 100, seed = seed),

message(sprintf("Model selected: %s | Split seed: %s", "LASSO", seed))

modelDesign <- PatientLevelPrediction::createModelDesign(
  targetId = 1,
  outcomeId = 2,
  populationSettings = populationSettings,
  covariateSettings = covariateSettings,
  preprocessSettings = preprocessSettings,
  modelSettings = modelSettings,
  splitSettings = splitSettings,
  runCovariateSummary = TRUE
)

plpModule <- PatientLevelPredictionModule$new()
plpModuleSpecs <- plpModule$createModuleSpecifications(modelDesignList = modelDesign)
```

:::

# Strategus Analysis Specifications

::: {.panel-tabset}

### Steps
With the cohort and PLP modules ready, assemble the Strategus analysis specifications and persist them for execution.

::: {.callout-note}
## Output Artifact
- [ ] `analysisSpecifications.json` captures the combined resources and module settings needed for execution.
:::

### R Code
```{r} 
#| label: analysis-specs

analysisSpecifications <- createEmptyAnalysisSpecifications() |>
  addSharedResources(cohortDefShared) |>
  addModuleSpecifications(cohortGeneratorModuleSpecifications) |>
  addModuleSpecifications(plpModuleSpecs)

ParallelLogger::saveSettingsToJson(analysisSpecifications, "analysisSpecifications.json")
```

:::

# Execution settings and running

::: {.panel-tabset}

### Steps
Create the execution settings, point Strategus to your DuckDB database, and then run the analysis.

::: {.callout-warning}
## Before You Run
- [ ] Confirm the work and results directories exist or can be created by Strategus.
- [ ] Make sure the DuckDB file path is correct and accessible.
- [ ] Expect execution to take a few minutes depending on data size.
:::

### R Code
```{r}
#| label: execute
workFolder <- normalizePath("./results/strategus/work")
resultsFolder <- normalizePath("./results/strategus/results")

executionSettings <- Strategus::createCdmExecutionSettings(
  workDatabaseSchema = workSchema,
  cdmDatabaseSchema = cdmSchema,
  cohortTableNames = CohortGenerator::getCohortTableNames(cohortTable = cohortTable),
  workFolder = workFolder,
  resultsFolder = resultsFolder,
  incremental = FALSE
)

connectionDetails <- DatabaseConnector::createConnectionDetails(
  dbms = "duckdb",
  server = dbPath
)

# Execute the study (may take some minutes depending on data size)
Strategus::execute(
  analysisSpecifications,
  executionSettings = executionSettings,
  connectionDetails = connectionDetails
)
```

:::

# View Results

::: {.callout-note collapse="true"}
## Interactive Review
Use the PLP viewer to inspect the results produced by the execution step. Expand this panel when you are ready to launch the shiny viewer.

```{r, eval=FALSE}
analysisLocation <- file.path(workFolder, "PatientLevelPredictionModule")
PatientLevelPrediction::viewMultiplePlp(analysisLocation)

```
:::
